% Appendix C

\chapter{Source Code} 
\label{AppendixC} 

All outputs of this thesis are stored in the GitHub \url{https://github.com/LasFo/ma-the}.

The structure is as follows:
\begin{itemize}
 \item \code{tests} contains the tests that were developed during the thesis. Most of thes test were
                    used to test for specific bugs during the development process. Thus some of these
                    test seem unnecessary. The subdirectory contains
                    tests that run indefinitely. 
 \item \code{implementation} contains the implementations STMWSL, STMLA and the Project implementation.
                             It does not contain the GHC implementation. Since it is no longer part of 
                             the standard libraries of GHC 8.0.1, it can be downloaded with \code{cabal}\footnote{\url{https://www.haskell.org/cabal/}}
 \item \code{recavoid} contains ideas and challenges for the problem of unnecessary recomputations.
 \item \code{stm} contains all code that were developed during the the first part of this thesis that is described in Appendix \ref{AppendixA}.
 \item \code{timing} contains the ressources that were used for the performance tests.
 \item \code{others} contains a documentation, this written part and slides for a talk.
\end{itemize}

The implementation was developed and tested with GHC 8.0.1. Other version were not tested. There 
is no guarantee that the implementation works for other versions. Serveral libraries are used 
within the implementation. All used libraries are part of the standard Distribution of GHC. 
The only library that is used but not in the standard Distribution is \code{Data.Tuple.Select}. 
The cabal package \code{tuple} includes this library. 