% Chapter 0

\chapter{Motivation} % Main chapter title

\label{Chapter0} % For referencing the chapter elsewhere, use \ref{Chapter1} 

%----------------------------------------------------------------------------------------

% Define some commands to keep the formatting separated from the content 
\newcommand{\keyword}[1]{\textit{#1}}
\newcommand{\tabhead}[1]{\textbf{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\file}[1]{\texttt{\bfseries#1}}
\newcommand{\option}[1]{\texttt{\itshape#1}}

Modern computer architecture includes multicore processors. To utilize these multicore system to their full extend, concurrent and parallel programming is needed.
By this new challenges arise. One challenge is the logical issue of splitting the problem in smaller problems which can be processed by different threads in parallel.
Aditionally there are technical challenges. For example a new scheduler is needed and hardware accesses (Printer, Display, etc.) need to be coordinated. These are challenges
the operating system usually handles. The are other challenges the operating system cannot handle, because they are specific for every program.

The most discussed challenge is the synchronization. If a program works with multiple threads, these threads usually communicate. Communications means to 
exchange data. Even a simple statement like an assignment can cause problem when used in the parallel threads. The problem is that these operations are
non atomic operations. Thus (\code{x = x + 1}) consist of three parts. first reading the old value, second adding \code{1}, and thrid write the new value.
This means two threads in parallel can both read the old value, then both add \code{1} to the old value, and then write back the new value. 
The new value is the initial value incremented by \code{1}, even though two threads executed an increment operation on this value. 
This non inteded behaviour is called \keyword{lost update}. The efforts to avoid non intended behaviour such as this are called synchronization.

Even though multicore processors are new, the research in the field of synchronization has a long history, starting with \parencite{semaphore}, which 
introduces the most basic synchronization tool, the semaphore. The semaphore is a abstract datatype which holds an Interger and provides two 
\keyword{atomic} operations, \code{P} and \code{V}. If the value of the semaphore is greater than \code{0}, \code{P} decrements the semaphore. If the value of 
the semaphore is \code{0} the thread that evoked \code{P} is suspended. When a thread evokes \code{V} the value of the semaphore is increased and
in the case another thread is currently suspended, because it called \code{P} on the semaphore, that thread is awakened. After the thread is
awakened, it tries \code{P} again. 

This seem to be a simple construct, but its capabilities are enormous. It is highly complex to use a semaphore correctly.
The main problem of semaphores is the so called deadlock\footnote{In the course of this thesis I will refer to deadlocks as a static propertie rather than a state of a system.}. 
This means there is a schedule, where no progress of the systen is possible, because all threads are waiting for a semaphore. The term deadlock is not exclusive for semaphores.
It is used for all blocking mechanisms. To avoid such deadlocks is very hard even when using one or few semaphores. 
It is nearly impossible to avoid deadlocks when you trie to compose semaphore based functions.

To avoid the problems of semaphores while maintaining the expressiveness of semaphores the so called software transactions were introduced \parencite{STMBase}.
Sofware transactions are inspired by the long known database transactions \parencite{DBTrans}. Software transactions provide an interface to program with 
single element buffers. If you are using this interface the underlying implementation ensures the so called \keyword{ACI(D)} properties. \textbf{A} for atomicity.
This means a transactions appears to be processed instantaneous. \textbf{C} for consistency. This means that a consistent view of the system is always guaranteed.
\textbf{I} stands for isolation. This means the programmer does not need to worry about concurrency and every thread can act as if it were the only thread. 
\textbf{D} stand for durability, but is relevant only for data base transactions. 

There is a stable implementation for software transactions in Haskell, namely Software Transactional Memory (called STM in the following). STM provides the ACI(D)
properties by optimistically executing the transaction. If a conflict is detected, the changes are discarded and the transactions is restarted (a so called rollback). 
This works, but is not optimal with regards to efficiency and performance. There are two problems. First the conflict detection. Sometimes the implementation detects 
a conflict and evokes a rollback, even though it is not necessary. The second problem is the rollback mechanism. Regardless of the conflict, always the whole transaction
is reexecuted. This includes operations on data that has not changed, thus an unnecessary recomputation. These problems are discussed in detail in Chapter \ref{Chapter1}. 
The aim of this thesis is to provide an alternative implementation that avoids these problems while preserving the ACI(D) properties.
